---
title: "Conclique-based Gibbs sampling"
author: "Andee Kaplan and Daniel Nordman"
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_vignette:
  fig_caption: yes
  toc: true
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Conclique-based Gibbs sampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Markov random field models are common for spatial data. Rather than specifying a joint distribution directly, a  model is specified through a set of full conditional distributions for each spatial location. The `R` package `conclique` provides a way to simulate from a MRF using a Gibbs sampler based on *concliques* as well as implements a formal goodness-of-fit (GOF) test for model assessment. For reference, see [@kaiser2012goodness].

This vignette shows an example of how to simulate from a MRF using a conclique-based Gibbs sampler using `conclique`. This example is extended to looking at GOF tests in [GOF statistics](gof-statistics.html)

#Concliques
*Concliques* are defined as sets of locations such that no location in the set is a neighbor
of any other location in the set. For example, the sets of all singular locations is a trivial set of concliques for any neighborhood structure. We can create the set of maximal concliques for a given lattice (the minimal conclique cover) using the `conclique` package and assign them to a grid for easy plotting.

```{r concliques}
library(conclique)

N <- 6
grid <- matrix(1:(N*N), nrow = N)

lattice <- lattice_4nn_torus(c(N, N))
concliques <- min_conclique_cover(lattice)
grid <- assign_concliques(grid = grid, conclique_cover = concliques)

```

```{r conc-plots, echo=FALSE, warning = FALSE, message = FALSE, fig.cap="Minimal conclique cover for a given lattice of size $6\\times6$ using a 4-nearest neighbors structure wrapped on a torus. This results in 2 concliques."}
library(ggplot2)
library(dplyr)
library(scales)
theme_blank <- function() {
  theme <- theme_bw()
  theme$line <- element_blank()
  theme$rect <- element_blank()
  theme$strip.text <- element_blank()
  theme$axis.text <- element_blank()
  theme$plot.title <- element_blank()
  theme$axis.title <- element_blank()
  
  return(theme)
}
ggplot2::theme_set(theme_blank())


set.seed(300)
expand.grid(x = 1:N, y = 1:N) %>%
  rowwise() %>%
  mutate(conclique = grid[x, y]) %>%
  ggplot() +
  geom_text(aes(x = x, y = y, label = conclique))
```

#Simulation
We can use the conditional independence of the concliques to create a batch updating Gibbs sampler to simulate spatial values from this model. The `conclique` package comes equipped with all the functions necessary to simulate from a Gaussian MRF with a single dependence parameter, however the user can supply his own functions for simulation, see the section on [extending](#extending-conclique) `conclique`. 

For this example, we will simulate from the Gaussian MRF with $\rho = 1, \kappa = 0, \eta = .24$ using both a conclique-based Gibbs sampler and a sequential Gibbs sampler to compare run time.

```{r}
neighbors <- get_neighbors(lattice)
params <- list(rho = 1, kappa = 0, eta = .24)
inits <- matrix(0, nrow = N, ncol = N)
num_sample <- 10000

## conclique-gibbs
time <- Sys.time()
conclique_simulation <- run_conclique_gibbs(concliques, neighbors, inits, "gaussian_single_param", params, num_sample)
conc_time <- Sys.time() - time 

## sequential gibbs
time <- Sys.time()
sequential_simulation <- run_sequential_gibbs(neighbors, inits, "gaussian_single_param", params, num_sample)
seq_time <- Sys.time() - time
```

In this example, the conclique-based gibbs sampler took `r conc_time` seconds and the sequential-based gibbs sampler took `r seq_time` seconds to simulate `r comma(num_sample)` spatial data sets of size $6 \times 6$.

#Extending `conclique`
Obviously, simulating from a Gaussian MRF with 4-nearest neighbor structure can be accomplished with exact simulation. However, the conclique-based Gibbs method can be used for an arbitrary dependence structure and a user-defined conditional distribution.

##Dependence structure
The `conclique` package comes with a function to create a lattice of specified dimension with 4-nearest neighbor structure wrapped on a torus (`lattice_4nn_torus`). However, the user can create his own dependence structure by creating an `igraph` [@igraph] object with each location as a node and the dependence as the edges. The dimension of the lattice must be kept as an attribute for the object, this is accomplished using

```
igraph::set.graph.attribute(lattice, "dimvector", dimvec)
```

where `dimvec` is a vector storing the dimensions of the lattice, for example `c(N, N)`.

##Conditional distribution
Additionally, the user can specify a difference conditional distribution to sample the spatial data from. In order to accomplish this, the user must provide a function which takes as parameters 

 - `data` A list containing two elements, `sums` and `nums`, which contain the sum of the data in each neighborhood as well as the number of locations in the neighborhood for each point in the conclique. 
 - `params` A named list of parameter values, that parameterize the distribution.
 
This function must return a sampled dataset the same size as the original data.

For example, the sampler function for the single dependency parameter Gaussian model is reproduced below.

```
gaussian_single_param_sampler <- function(data, params) {
  rho <- params$rho
  kappa <- params$kappa
  eta <- params$eta
  
  mean_structure <- kappa + eta*(data$sums - data$nums*kappa)
  rnorm(length(mean_structure))*rho + mean_structure
}
```

#References