---
title: "GOF Statistics"
author: "Andee Kaplan and Daniel Nordman"
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_vignette:
  fig_caption: yes
  toc: true
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{GOF Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Markov random field models are common for spatial data. Rather than specifying a joint distribution directly, a  model is specified through a set of full conditional distributions for each spatial location. The `R` package `conclique` provides a way to simulate from a MRF using a Gibbs sampler based on *concliques* as well as implements a formal goodness-of-fit (GOF) test for model assessment. For reference, see [@kaiser2012goodness].

This vignette shows an example of how to test for Goodness-of-Fit (GOF) using a conclique-based approach in the package `conclique`. This example is preceeded by an example of simulating spatial data using conclique-based Gibbs sampling in the vignette [Conclique-based Gibbs sampling](conclique-gibbs.html)

```{r concliques-contd, echo=FALSE, message = FALSE, warning = FALSE}
library(conclique)
library(dplyr)
library(tidyr)
library(ggplot2)

set.seed(300)
N <- 6
grid <- matrix(1:(N*N), nrow = N)

lattice <- lattice_4nn_torus(c(N, N))
concliques <- min_conclique_cover(lattice)

neighbors <- get_neighbors(lattice)
params <- list(rho = 1, kappa = 0, eta = .24)
inits <- matrix(0, nrow = N, ncol = N)
num_sample <- 10000

conclique_simulation <- run_conclique_gibbs(lattice, 
                                            concliques, 
                                            neighbors, 
                                            inits, 
                                            gaussian_single_param_sampler, 
                                            params, 
                                            n.iter = num_sample)

sequential_simulation <- run_sequential_gibbs(lattice, 
                                            neighbors, 
                                            inits, 
                                            gaussian_single_param_sampler, 
                                            params, 
                                            n.iter = num_sample)

```

#Generalized Spatial Residuals
Let $F(y|\boldsymbol y(N_i), \boldsymbol \theta)$ be the conditional cdf of the data, $Y(\boldsymbol s_i)$ under the model. We can define *generlized spatial residuals* through substitution of random variables, $Y(\boldsymbol s_i)$ and neighbors $\{Y(\boldsymbol s_j): \boldsymbol s_j \in N_i \}$, into  (continuous) conditional cdf:

$$
R(\boldsymbol s_i) = F(Y(\boldsymbol s_i)|\{Y(\boldsymbol s_j): \boldsymbol s_j \in N_i \}, \boldsymbol \theta).
$$

These residuals, $R(\boldsymbol s_i)$, can be defined for non-continuous $F$ too.

It then holds that *within* a conclique, the generalized spatial residuals are iid Uniform$(0, 1)$-distributed.

The spatial residuals are obtainable using the `conclique` package. We can continue with our example from the vignette [Conclique-based Gibbs sampling](conclique-gibbs.html) using data generated from a Gaussian MRF with $\rho = 1, \kappa = 0, \eta = .24$. Use the last dataset we generated as our "data".

```{r resids}
conclique_dat <- conclique_simulation[num_sample, ]
conclique_resids <- spatial_residuals(conclique_dat, neighbors, gaussian_single_param_cdf, params)
```

```{r resids-plot, echo=FALSE, fig.cap="Looking at the emprirical cdf of generalized spatial residuals from simulated Gaussian MRF with $\\rho = 1, \\kappa = 0, \\eta = .24$ versus the correct model. The black line represents the cdf of a Uniform$(0,1)$."}
ecdf_vals <- lapply(concliques, function(conc) {fn <- ecdf(conclique_dat[conc]); fn(conclique_dat[conc])})
names(ecdf_vals) <- 1:length(concliques)
ecdf_vals <- do.call(cbind, ecdf_vals) %>% data.frame()

ecdf_vals %>%
  data.frame() %>%
  gather(conclique, ecdf) %>%
  separate(conclique, into = c("junk", "conclique"), 1) -> ecdf_vals

ecdf_vals[, "u"] <- do.call(c, lapply(concliques, function(conc) conclique_resids[conc]))

ecdf_vals %>%
  ggplot() +
  geom_point(aes(u, ecdf, colour = conclique)) +
  geom_abline(aes(slope = 1, intercept = 0)) +
  theme_bw() +
  theme(aspect.ratio = 1)

```

We could also look at the residuals from a poorly fit model, say with $\eta = 0.05$.

```{r resids_err}
conclique_resids_err <- spatial_residuals(conclique_dat, neighbors, gaussian_single_param_cdf, list(rho = 1, kappa = 0, eta = 0.05))
```

```{r resids_err-plot, echo=FALSE, fig.cap="Looking at the emprirical cdf of generalized spatial residuals from simulated Gaussian MRF with $\\rho = 1, \\kappa = 0, \\eta = .24$ versus an incorrect model with $\\eta = .05$. The black line represents the cdf of a Uniform$(0,1)$."}
ecdf_vals_err <- lapply(concliques, function(conc) {fn <- ecdf(conclique_dat[conc]); fn(conclique_dat[conc])})
names(ecdf_vals_err) <- 1:length(concliques)
ecdf_vals_err <- do.call(cbind, ecdf_vals_err) %>% data.frame()

ecdf_vals_err %>%
  data.frame() %>%
  gather(conclique, ecdf) %>%
  separate(conclique, into = c("junk", "conclique"), 1) -> ecdf_vals_err

ecdf_vals_err[, "u"] <- do.call(c, lapply(concliques, function(conc) conclique_resids_err[conc]))

ecdf_vals_err %>%
  ggplot() +
  geom_point(aes(u, ecdf, colour = conclique)) +
  geom_abline(aes(slope = 1, intercept = 0)) +
  theme_bw() +
  theme(aspect.ratio = 1)

```
We can see that for both concliques, all of the residuals are below the Uniform$(0,1)$ cdf, showing this may be a poor fit (which in fact it is by design).

#Test Statistics
In order to combine generalized spatial residuals across concliques to form a single test statistic, we look at the empirical cdf of the residuals and its difference to the Uniform$(0,1)$ cdf using some discrepancy measure (like the Kolmogorov-Smirnov statistic) and combine using an aggregation function (like maximum).

Both the Kolmogorov-Smirnov statistic and the Cramèr-von-Mises criterion are implemented in `conclique`, and any user specified aggregation can be supplied; `mean` and `max` are recommended.

```{r gof}
conclique_gof <- gof_statistics(conclique_resids, concliques, "ks", "max")
conclique_gof_err <- gof_statistics(conclique_resids_err, concliques, "ks", "max")
```

```{r gof_table, echo=FALSE}
data.frame(True = conclique_gof, Incorrect = conclique_gof_err) %>%
  knitr::kable()

```

We can see that the incorrectly specified model has a higher value of the statistic, which points to potentially rejecting this model, however without a reference distribution, it's impossible to say. Luckily, we have the ability to sample from the model and bootstrap this distribution using our simulation functions in `conclique`.

#Extending `conclique`
Once again, one of the key advantages to using conclique-based approaches for simulation and GOF tests is the ability to extend to models beyond the Gaussian MRF with 4-nearest neighbor structure. In order to extend `conclique` to use an arbitrary conditional distibution for the generalized spatial residuals, the user must specify a cdf function which takes as parameters 

 - `data` A list containing two elements, `sums` and `nums`, which contain the sum of the data in each neighborhood as well as the number of locations in the neighborhood for each point in the conclique. 
 - `params` A named list of parameter values, that parameterize the distribution.
 
For example, the cdf function for the single dependency parameter Gaussian model is reproduced below.

```
gaussian_single_param_cdf <- function(data, params) {
  rho <- params$rho
  kappa <- params$kappa
  eta <- params$eta
  
  mean_structure <- kappa + eta*(data$sums - data$nums*kappa)
  pnorm(data$data, mean = mean_structure, sd = rho)
}
```

#References
